quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-10-16 paper-quollr.Rmd
Adde file structures into Usage
Added initial working for the implementation

To begin the workflow, users need two inputs: the high-dimensional dataset and the corresponding nonlinear dimensionality reduction (NLDR) layout. The high-dimensional data must contain a unique `ID` column, with data columns prefixed by the letter `"x"` (e.g., `x1`, `x2`, etc.). The NLDR dataset should include embedding coordinates labeled as `emb1` and `emb2`, ensuring one-to-one correspondence with the high-dimensional data through the shared `ID`.

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. To illustrate the workflow, we use the built-in example dataset `scurve`, a $7\text{-}D$ simulated dataset consisting of $1000$ observations. The first three variables define a $3\text{-}D$ S-shaped manifold, while the remaining four variables introduce low-magnitude uniform noise, yielding a structured yet noisy high-dimensional dataset.

>>>>

The implementation of `quollr` emphasizes modularity, computational efficiency, and interoperability.
Its design follows a layered architecture that separates **data preprocessing**, **model construction**, and **visualization**, ensuring clarity, extensibility, and compatibility with the wider R ecosystem.

## Software architecture

Internally, `quollr` is organized into five primary modules that correspond to key steps in the analysis pipeline:

1. **Data preparation** — functions such as `scale_highd()` and `prepare_data()` standardize the input, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
2. **Hexagonal binning** — the `hex_bin_data()` module constructs a two-dimensional hexagonal grid and aggregates local summaries of the high-dimensional data for each bin.
3. **Model fitting** — the `fit_highd_model()` function computes high-dimensional centroids and pairwise bin connections to build a structured manifold representation.
4. **Error computation** — functions like `compute_errors()` and `aggregate_bin_error()` evaluate local and global discrepancies between the model and data, enabling diagnostic visualization.
5. **Visualization and linkage** — plotting and tour functions (`show_hex_model()`, `show_link_plots()`, and `show_error_link_plots()`) integrate outputs into linked static and dynamic displays using `ggplot2`, `langevitour`, and `crosstalk`.

Each module is internally independent but communicates through standardized data objects (see next section).
This modular design simplifies maintenance and allows developers to extend individual components—such as substituting different binning strategies, embedding methods, or visualization tools—without altering the overall workflow.

## Data objects

The internal data objects in `quollr` adhere to the **tidy data** principle, ensuring that every variable forms a column, every observation forms a row, and every type of observational unit forms a table.
This structure supports compatibility with the `tidyverse` and other R visualization frameworks.

The main object classes include:

* **`highd_data`** – a tibble containing the original high-dimensional observations with a unique identifier (`ID`) and variable columns prefixed with `"x"` (e.g., `x1`, `x2`, …).
* **`nldr_data`** – a tibble containing two-dimensional embeddings, labeled as `emb1` and `emb2`, matched to the same `ID`s.
* **`highd_vis_model`** – the primary model object returned by `fit_highd_model()`, containing two key components: the high-dimensional summaries for each bin and their corresponding two-dimensional positions.
* **`hex_bin_obj`** – the internal representation of the hexagonal grid, including bin coordinates (`a1`, `a2`), centroids, and normalized weights (`w_h`).

These objects provide a coherent foundation for data transfer across `quollr`’s modules—from binning to model fitting and visualization—supporting reproducibility and transparency in analyses.

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. To begin the workflow, users need two inputs: the high-dimensional dataset and the corresponding nonlinear dimensionality reduction (NLDR) layout. The high-dimensional data must contain a unique `ID` column, with data columns prefixed by the letter `"x"` (e.g., `x1`, `x2`, etc.). The NLDR dataset should include embedding coordinates labeled as `emb1` and `emb2`, ensuring one-to-one correspondence with the high-dimensional data through the shared `ID`.

To illustrate the workflow, we use the built-in example dataset `scurve`, a $7\text{-}D$ simulated dataset consisting of $1000$ observations. The first three variables define a $3\text{-}D$ S-shaped manifold, while the remaining four variables introduce low-magnitude uniform noise, yielding a structured yet noisy high-dimensional dataset.

### 2025-10-16 paper-quollr.tex
Adde file structures into Usage
Added initial working for the implementation

To begin the workflow, users need two inputs: the high-dimensional dataset and the corresponding nonlinear dimensionality reduction (NLDR) layout. The high-dimensional data must contain a unique \texttt{ID} column, with data columns prefixed by the letter \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, etc.). The NLDR dataset should include embedding coordinates labeled as \texttt{emb1} and \texttt{emb2}, ensuring one-to-one correspondence with the high-dimensional data through the shared \texttt{ID}.

@@

The following demonstration of the package's functionality assumes \texttt{quollr} has been loaded. To illustrate the workflow, we use the built-in example dataset \texttt{scurve}, a \(7\text{-}D\) simulated dataset consisting of \(1000\) observations. The first three variables define a \(3\text{-}D\) S-shaped manifold, while the remaining four variables introduce low-magnitude uniform noise, yielding a structured yet noisy high-dimensional dataset.

>>>>

The implementation of \texttt{quollr} emphasizes modularity, computational efficiency, and interoperability.
Its design follows a layered architecture that separates \textbf{data preprocessing}, \textbf{model construction}, and \textbf{visualization}, ensuring clarity, extensibility, and compatibility with the wider R ecosystem.

\subsection{Software architecture}\label{software-architecture}

Internally, \texttt{quollr} is organized into five primary modules that correspond to key steps in the analysis pipeline:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Data preparation} --- functions such as \texttt{scale\_highd()} and \texttt{prepare\_data()} standardize the input, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
\item
  \textbf{Hexagonal binning} --- the \texttt{hex\_bin\_data()} module constructs a two-dimensional hexagonal grid and aggregates local summaries of the high-dimensional data for each bin.
\item
  \textbf{Model fitting} --- the \texttt{fit\_highd\_model()} function computes high-dimensional centroids and pairwise bin connections to build a structured manifold representation.
\item
  \textbf{Error computation} --- functions like \texttt{compute\_errors()} and \texttt{aggregate\_bin\_error()} evaluate local and global discrepancies between the model and data, enabling diagnostic visualization.
\item
  \textbf{Visualization and linkage} --- plotting and tour functions (\texttt{show\_hex\_model()}, \texttt{show\_link\_plots()}, and \texttt{show\_error\_link\_plots()}) integrate outputs into linked static and dynamic displays using \texttt{ggplot2}, \texttt{langevitour}, and \texttt{crosstalk}.
\end{enumerate}

Each module is internally independent but communicates through standardized data objects (see next section).
This modular design simplifies maintenance and allows developers to extend individual components---such as substituting different binning strategies, embedding methods, or visualization tools---without altering the overall workflow.

\subsection{Data objects}\label{data-objects}

The internal data objects in \texttt{quollr} adhere to the \textbf{tidy data} principle, ensuring that every variable forms a column, every observation forms a row, and every type of observational unit forms a table.
This structure supports compatibility with the \texttt{tidyverse} and other R visualization frameworks.

The main object classes include:

\begin{itemize}
\tightlist
\item
  \textbf{\texttt{highd\_data}} -- a tibble containing the original high-dimensional observations with a unique identifier (\texttt{ID}) and variable columns prefixed with \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, \ldots).
\item
  \textbf{\texttt{nldr\_data}} -- a tibble containing two-dimensional embeddings, labeled as \texttt{emb1} and \texttt{emb2}, matched to the same \texttt{ID}s.
\item
  \textbf{\texttt{highd\_vis\_model}} -- the primary model object returned by \texttt{fit\_highd\_model()}, containing two key components: the high-dimensional summaries for each bin and their corresponding two-dimensional positions.
\item
  \textbf{\texttt{hex\_bin\_obj}} -- the internal representation of the hexagonal grid, including bin coordinates (\texttt{a1}, \texttt{a2}), centroids, and normalized weights (\texttt{w\_h}).
\end{itemize}

These objects provide a coherent foundation for data transfer across \texttt{quollr}'s modules---from binning to model fitting and visualization---supporting reproducibility and transparency in analyses.

@@

The following demonstration of the package's functionality assumes \texttt{quollr} has been loaded. To begin the workflow, users need two inputs: the high-dimensional dataset and the corresponding nonlinear dimensionality reduction (NLDR) layout. The high-dimensional data must contain a unique \texttt{ID} column, with data columns prefixed by the letter \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, etc.). The NLDR dataset should include embedding coordinates labeled as \texttt{emb1} and \texttt{emb2}, ensuring one-to-one correspondence with the high-dimensional data through the shared \texttt{ID}.

To illustrate the workflow, we use the built-in example dataset \texttt{scurve}, a \(7\text{-}D\) simulated dataset consisting of \(1000\) observations. The first three variables define a \(3\text{-}D\) S-shaped manifold, while the remaining four variables introduce low-magnitude uniform noise, yielding a structured yet noisy high-dimensional dataset.
